---
lecture: "13a"
metadata-files:
  - ../../_slide-settings.yml
format:
  revealjs:
    theme: default
---

::: {.my-title}
# [Multilevel Modeling]{.blue}
Troubleshooting

::: {.my-grey}
[{{< meta semester >}} | CLAS | {{< meta course >}}]{}<br />
[Jeffrey M. Girard | Lecture {{< meta lecture >}}]{}
:::

![](../../img/city-girl.svg){.absolute bottom=30 right=0 width="400"}
:::

```{r setup}
#| echo: false
#| message: false

library(tidyverse)
library(easystats)
library(patchwork)
library(here)
library(glmmTMB)

source(here("_common.R"))

set_theme(theme_classic(base_size = 20))

options(width = 75)
```

## Roadmap

::: {.columns .pv4}

::: {.column width="60%"}
1. Understanding Non-Convergence
    + *The Likelihood Surface*
    + *Diagnosing Warnings*
    
2. A Systematic Workflow
    + *Top-Down Troubleshooting*
    + *Bayesian Estimation*
    
:::

::: {.column .tc .pv4 width="40%"}
{{< lif "../../icons/map.json" trigger=hover colors=secondary:#2a76dd class=rc >}}
:::

:::

# Troubleshooting

## Convergence Problems {.fsmaller}

:::{.columns}
:::{.column width="60%"}
- MLMs are inherently complex and datasets often lack the information needed for extreme parameterization.
- Convergence problems are a routine part of the MLM workflow.
- [Do not ignore these warnings!]{.b .dark-red} Parameter estimates from non-converged models are mathematically unreliable.
- However, do not despair. Troubleshooting is a systematic process and there are standard steps to fix these issues.
:::
:::{.column .tc .pv4 width="40%"}
{{< lif "../../icons/ghost.json" trigger=hover colors=secondary:#2a76dd class=rc >}}
:::
:::

## The Likelihood Surface {.fsmaller}

Before diving into specific warnings, we must remember back to Lecture 04b.

- **Maximum Likelihood Estimation:** The optimizer searches for a combination of parameter values that maximizes the probability of observing the data.
- **The Mountain Analogy:** Imagine a multi-dimensional mountain. The optimizer takes steps to find the absolute highest peak.
- **The Map and Compass:** The optimizer uses local derivative information (the Gradient and Hessian) to aid in the search.
- **The Problem:** If the mountain has a flat ridge, multiple peaks, or a boundary it cannot cross, the optimizer gets confused and either throws a warning or silently stops at an invalid parameter value.

## Optimization as Hill Climbing {.fsmaller}

a. Select a random or "best guess" starting value^[We can optionally use a "multi-start" strategy (for multiple climbers)]
b. Estimate the likelihood of the current value^[Technically, we try to minimize the negative log-likelihood]
c. Detemine whether to increase or decrease the value^[We can optionally use derivative information (gradient/hessian) here]
d. Update the current value by some amount (step size)^[We can adjust the step size, which affects speed and pathing]
e. Repeat steps 2-4 until the change in likelihood is small^[We repeat until convergence or a maximum number of iterations]

## Visual Example 

```{r}
#| echo: false

likelihood_function <- function(x) {
  -0.1 * (x - 5)^2 + 
    3 * exp(-((x - 2)^2) / 0.5) +  # Local peak at x = 2
    2 * exp(-((x - 8)^2) / 0.8) +  # Local peak at x = 8
    3
}

x_vals <- seq(0, 10, length.out = 101)

likelihood_vals <- likelihood_function(x_vals)

df <- tibble(estimate = x_vals, likelihood = likelihood_vals)

ggplot() +
  geom_line(
    data = filter(df, estimate <= 2),
    aes(estimate, likelihood),
    linewidth = 1.5, 
    color = "steelblue3"
  ) +
  geom_line(
    data = filter(df, estimate >= 8),
    aes(estimate, likelihood),
    linewidth = 1.5, 
    color = "coral3"
  ) +
  geom_point(
    data = tibble(x = 2, y = likelihood_vals[x_vals == 2]),
    aes(x, y),
    color = "steelblue3",
    size = 5
  ) +
  geom_point(
    data = filter(df, estimate %in% c(0, 5, 10)),
    aes(estimate, likelihood),
    size = 6, 
    color = "black"
  ) +
  geom_point(
    data = tibble(x = 5, y = likelihood_vals[x_vals == 5]),
    aes(x, y),
    color = "seagreen",
    size = 5
  ) +
  geom_point(
    data = tibble(x = 8, y = likelihood_vals[x_vals == 8]),
    aes(x, y),
    color = "coral3",
    size = 5
  ) +
  geom_segment(
    data = tibble(x = -1, xend = 2, y = likelihood_vals[x_vals == 2]),
    aes(x = x, xend = xend, y = y, yend = y),
    color = "steelblue3",
    linewidth = 1,
    linetype = "dashed"
  ) +
  geom_segment(
    data = tibble(x = -1, xend = 5, y = likelihood_vals[x_vals == 5]),
    aes(x = x, xend = xend, y = y, yend = y),
    color = "seagreen",
    linewidth = 1,
    linetype = "dashed"
  ) +
  geom_segment(
    data = tibble(x = -1, xend = 8, y = likelihood_vals[x_vals == 8]),
    aes(x = x, xend = xend, y = y, yend = y),
    color = "coral3",
    linewidth = 1,
    linetype = "dashed"
  ) +
  labs(x = "Parameter Estimate", y = "Likelihood") +
  coord_cartesian(ylim = c(0, 6), xlim = c(0, 10)) +
  theme_bw(base_size = 36)
```

## Setup 

```{r}
#| message: false

library(tidyverse)
library(glmmTMB)
library(easystats)
```

```{r}
#| message: false
#| replace_path: ["../../data/", ""]

trouble <- read_csv("../../data/trouble_sim.csv")
glimpse(trouble)
```

# Stage 1:<br> Data Fixes

## Problem 1: Extreme Scaling

:::{.fsmaller}
A frequent warning in `glmmTMB` is a "non-positive-definite Hessian matrix." This simply means the optimizer cannot mathematically verify it has found a peak. One common cause is a numerical issue resulting from variables living on vastly different mathematical scales.
:::

```{r}
#| warning: true
#| message: true

# Try to predict a massive y from a tiny x
fit_scaling <- glmmTMB(
  formula = viral_load ~ 1 + dosage_kg + (1 | clinic_id), 
  data = trouble
)
```

## Detecting Extreme Scaling

:::{.fsmaller}
Look for means/medians or ranges that are on wildly different scales
:::

```{r}
trouble |> select(viral_load, dosage_kg) |> summary()
```


## Addressing Extreme Scaling

::: {.fsmaller}
**The Fix:** We need to get the variables on similar scales. It is recommended to *rescale* outcomes (to preserve ICC) and *standardize* predictors. Here, we can divide `viral_load` by 10,000 to get its mean down from \~1M to \~100.
:::

```{r}
# Fix scaling issues
trouble_fix <- trouble |> 
  mutate(
    viral_load = viral_load / 10000,   # y = rescale
    dosage_kg = standardize(dosage_kg) # x = standardize
  )
```

:::{.fragment}
```{r}
# Refit and check
fit_scaling_fixed <- glmmTMB(
  formula = viral_load ~ 1 + dosage_kg + (1 | clinic_id), 
  data = trouble_fix
)
check_convergence(fit_scaling_fixed)
```
:::

## Problem 2: Collinearity

:::{.fsmaller}
The Hessian warning can also be triggered when the optimizer gets stuck on a flat likelihood ridge. A second major cause for this is severe collinearity among predictors, which confuses the optimizer because multiple parameter combinations yield the same model fit.
:::

```{r}
#| warning: true
#| message: true

# Try to predict y from two highly correlated x variables 
fit_collin <- glmmTMB(
  formula = cortisol_level ~ 1 + stress_score + anxiety_score + 
    (1 + stress_score + anxiety_score | clinic_id), 
  data = trouble
)
```

## Verifying Collinearity

::: {.fsmaller}
We need to remove the redundant information. It is recommended to check Variance Inflation Factors (VIFs) and then drop one of the collinear predictors. Recall that VIFs of 5+ indicate *moderate* and 10+ indicate *high* collinearity.
:::

```{r}
#| warning: true
#| message: true

# NB: I am setting ci = NA only to save space on slide
check_collinearity(fit_collin, ci = NA)
```

## Addressing Collinearity

```{r}
#| warning: true
#| message: true

# Drop anxiety_score, refit, and check
fit_collin_fixed <- glmmTMB(
  formula = cortisol_level ~ 1 + stress_score + 
    (1 + stress_score | clinic_id), 
  data = trouble
)
check_convergence(fit_collin_fixed)
```

# Stage 2:<br> Structural Fixes

## Problem 3: Overparameterization

:::{.fsmaller}
A third cause of the Hessian warning is when the optimizer hits a mathematical boundary. However, note that `glmmTMB` will sometimes converge silently despite this issue! This occurs when a variance is estimated near 0 (e.g., e-13) or a correlation between random effects is near +/- 1.00. The problem is that the model is too complex for the data.
:::

```{r}
#| warning: true
#| message: true

fit_overparam <- glmmTMB(
  formula = blood_pressure ~ 1 + age_c + (1 + age_c | clinic_id), 
  data = trouble
)
```

## Identifying Overparameterization

::: {.fsmaller}
To figure out if overparameterization caused the warning, you must actively inspect your random effect parameters.
:::

```{r}
# Inspect random effect parameters
model_parameters(fit_overparam, effects = "random")
```

::: {.fsmaller}
Notice the "smoking gun" in the output: the random effect correlation is nearly --1 and the random slope standard deviation is essentially zero.
:::

## Step 1: Remove Covariances

::: {.fsmaller}
We start by simplifying the structure without dropping variables. We force the correlation between the intercept and slope to be zero using `diag()`.
:::

:::{.fragment}
```{r}
fit_nocov <- glmmTMB(
  formula = blood_pressure ~ 1 + age_c + diag(1 + age_c | clinic_id), 
  data = trouble
)
model_parameters(fit_nocov, effects = "random")
```
:::

## The Zero-Correlation Assumption

::: {.fsmaller}
Using `diag()` is a significant change to your model. It explicitly assumes that the correlations between your random effects are zero (e.g., $\tau_{01}=0$).

- **Pooling Information:** By allowing $r \neq 0$, the model "pools information" across parameters. Knowing a cluster has a high intercept helps the model estimate its slope. `diag()` places parameters in "silos," preventing the model from "borrowing strength" from these relationships.
- **Substantive Trade-off:** You may gain model stability but you lose the ability to estimate and interpret the RE correlation(s).
- **Statistical Trade-off:** If strong correlation(s) exist, forcing them to zero can bias SEs, though this may be preferable to non-convergence.
:::

## Step 2: Prune Random Slopes

::: {.fsmaller}
In our updated results, the random slope SD was still very close to zero even after setting the RE correlation to zero. The next step may be to drop it (i.e., estimate a fixed slope only, assuming it is the same across all clusters).
:::

```{r}
fit_reduced <- glmmTMB(
  formula = blood_pressure ~ 1 + age_c + (1 | clinic_id), 
  data = trouble
)
model_parameters(fit_reduced, effects = "random")
```

## The Pruning Dilemma

::: {.fsmaller}
While we simplify to reach convergence, we must be careful with our tests.
:::

> **CRITICAL RULE:** Be extremely cautious when dropping the random slope for the specific fixed-effect predictor you are hypothesis testing.

::: {.fsmaller}
- **The Risk:** If a true random slope exists but you drop it to force a struggling model to converge, you inflate your Type I error rate. A $p < .05$ in a model with a missing focal random slope is often a false positive.
- **The Exception:** If the variance is estimated at zero, dropping the slope does not inflate Type I error because the model wasn't using it anyway.
:::

# Stage 3:<br> Algorithmic Fixes

## Tweak the Algorithm {.fsmaller}

If your variables are scaled, not perfectly collinear, and your structure is justified, but `glmmTMB` still throws a warning, the algorithm may just need more time or a different search strategy.

:::{.fragment}
**Increase the number of iterations and evaluations:**
```r
fit_algo <- glmmTMB(
  formula = ..., data = ...,
  control = glmmTMBControl(
    optCtrl = list(iter.max = 1e5, eval.max = 1e5)
  )
)
```
:::

:::{.fragment}
**Change the optimizer (e.g., to standard optim using BFGS):**
```r
fit_algo <- glmmTMB(
  formula = ..., data = ...,
  control = glmmTMBControl(
    optimizer = optim, 
    optArgs = list(method = "BFGS")
  )
)
```
:::

## Troubleshooting Cheatsheet {.fsmaller}

When your model throws a convergence warning (or produces boundary estimates), follow this general order of operations:

1.  **Stage 1: Fix your Data.** Check for extreme scaling or severe collinearity. 
    * *Fix:* Center, scale, or remove redundant variables.
2.  **Stage 2: Fix your Structure (Top-Down).** Did a variance hit near zero or a correlation hit near $\pm1.00$?
    * *Step A:* Remove RE covariances (e.g., using `diag()`).
    * *Step B:* Prune zero-variance components.
    * *Rule:* Protect the focal random slope unless its variance is truly zero.
3.  **Stage 3: Tweak the Algorithm.** If the model is fully justified but struggling, increase iterations or change the optimizer via `glmmTMBControl()`.

# Bayes

## Bayesian Estimation {.fsmaller}

Maximum Likelihood Estimation searches for a single highest peak. When the data lacks information, that peak disappears or becomes impossible to find.

:::{.fragment}
- **The Bayesian Solution:** Instead of finding a single point, Bayesian estimation maps out the entire landscape of probable values.
- **Priors as Guardrails:** By providing prior information, we can prevent the model from wandering into absurd territory (e.g., negative variances).
- Two popular R packages for Bayesian MLM are \{brms\} and \{rstanarm\}. I personally prefer the former, but the latter is easier to start with.
- \{rstanarm\} can be be installed from CRAN and chooses reasonable weak priors for you.
- Both use Stan to perform Markov Chain Monte Carlo (MCMC). 
:::

## MCMC Overview {.fsmaller}

- **Exploring vs. Climbing:** If MLE is a focused climber racing to find the single highest peak, MCMC is a wanderer exploring the entire mountain range to map its total volume and shape.

- **Volume over Altitude:** MLE fails when a peak is flat or hidden; MCMC succeeds by spending time proportionally: it stays longer in high-probability areas but intentionally visits lower-probability valleys to capture the full range of uncertainty.

- **Sampling over Solving:** Instead of using a compass (derivatives) to find the fastest path "up", MCMC uses a stochastic "random walk" to build a massive collection of representative snapshots of the landscape.

- **The Result:** MLE ends at a single coordinate (the "Best Guess"); MCMC results in a "Posterior Distribution" of thousands of valid guesses that reveal exactly how much we should trust our estimate.

## Example syntax of rstanarm

```{r}
#| message: false
#| cache: true
#| label: bfit

library(rstanarm)

# Problem 3: Boundary Issue on Random Slope SD
bfit <- stan_lmer(
  formula = blood_pressure ~ 1 + age_c + (1 + age_c | clinic_id), 
  data = trouble,
  chains = 4, # how many explorers to send out in total?
  cores = 4   # how many explorers to run simultaneously?
)
```

## Example output of rstanarm {.fsmaller}

```{r}
model_parameters(bfit, effects = "all")
```

- Median and 95% CI are point and interval estimates
- pd is probability of direction (significant if > 97.5%)
- Converged if all Rhat < 1.1 and ESS > 400 (i.e., chains &times; 100)
