---
lecture: "13b"
metadata-files:
  - ../../_slide-settings.yml
format:
  revealjs:
    theme: default
---

::: {.my-title}
# [Multilevel Modeling]{.blue}
Model Diagnostics

::: {.my-grey}
[{{< meta semester >}} | CLAS | {{< meta course >}}]{}<br />
[Jeffrey M. Girard | Lecture {{< meta lecture >}}]{}
:::

![](../../img/city-girl.svg){.absolute bottom=30 right=0 width="400"}
:::

```{r setup}
#| echo: false
#| message: false

library(tidyverse)
library(easystats)
library(patchwork)
library(here)
library(glmmTMB)

source(here("_common.R"))

set_theme(theme_classic(base_size = 20))

options(width = 75)
```

## Roadmap

::: {.columns .pv4}

::: {.column width="60%"}
1. A Systematic Workflow
    + *Predictive Performance*
    + *Visual Diagnostics*

2. Sensitivity & Easy Fixes
    + *Bootstrapping & LSM*
    + *DHARMa & Outliers*
    
:::

::: {.column .tc .pv4 width="40%"}
{{< lif "../../icons/map.json" trigger=hover colors=secondary:#2a76dd class=rc >}}
:::

:::

# Part 1:<br> The Philosophy

## Why Diagnostics?

- MLMs are complex. We aren't just estimating a line; we are estimating a **system of variances**.
- Our standard errors and $p$-values assume this system is perfectly Specified, Normal, and Constant.
- Small violations often don't matter, but...<br> **systemic violations** invalidate your results.

## Pass/Fail vs. Sensitivity

- It is rare to see "perfect" diagnostic plots in real research.
- Instead of asking "Did I pass?", ask:<br>**"How robust are my conclusions?"**
- If your finding disappears as soon as you account for a violation, the finding was likely an artifact of a bad model.

## The Mathematical Foundation

Recall the equations we use to define the random components of our models. They list many of our assumptions.

:::{.tc .b}
Level One (Observation)
:::

$$
\color{#4daf4a}{e_{ij}} \sim \text{Normal}(0, \color{#4daf4a}{\sigma})
$$

:::{.tc .b}
Level Two (Cluster)
:::

$$
\begin{bmatrix}\color{#e41a1c}{u_{0j}}\\ \color{#e41a1c}{u_{1j}}\end{bmatrix} \sim \text{MVN}\left(\begin{bmatrix}0\\0\end{bmatrix}, \begin{bmatrix}\color{#e41a1c}{\tau_{00}} & \\ \color{#e41a1c}{\tau_{10}} & \color{#e41a1c}{\tau_{11}}\end{bmatrix}\right)
$$

## Assumption 1: Specification

- All relevant predictors are included at the correct levels.
- The random effect structure ($\color{#e41a1c}{u_j}$) matches the data design.
- **The Random Slope Rule**:<br>If an effect varies across clusters, omitting the random slope forces that variation into the observation error ($\color{#4daf4a}{e_{ij}}$), which violates the independence of errors.

## Assumption 2: Normality

The equations explicitly declare the distribution shape:

- $\color{#4daf4a}{e_{ij}} \sim \textbf{Normal}(0, \color{#4daf4a}{\sigma}) \therefore\text{Univariate Normal}$
- $\color{#e41a1c}{u_{j}} \sim \textbf{MVN}(0, \color{#e41a1c}{\Sigma}) \therefore\text{Multivariate Normal}$

We assume Level 1 errors and Level 2 deviations follow a normal curve. If the real errors are non-normal, the likelihood math "breaks" and gives biased confidence intervals.

## Assumption 3: Homoscedasticity

Homoscedasticity means "constant variance."

:::{.fs80}
$$
\color{#4daf4a}{e_{ij}} \sim \text{Normal}(0, \color{#4daf4a}{\sigma}) \qquad \begin{bmatrix}\color{#e41a1c}{u_{0j}}\\ \color{#e41a1c}{u_{1j}}\end{bmatrix} \sim \text{MVN}\left(\begin{bmatrix}0\\0\end{bmatrix}, \begin{bmatrix}\color{#e41a1c}{\tau_{00}} & \\ \color{#e41a1c}{\tau_{10}} & \color{#e41a1c}{\tau_{11}}\end{bmatrix}\right)
$$
:::

- The model specifies exactly **one** $\color{#4daf4a}{\sigma}$ for all $i$ and $j$. It assumes the spread (SD) of errors is identical for every observation.
- The model specifies exactly **one** $\color{#e41a1c}{\tau_{00}}$ and $\color{#e41a1c}{\tau_{11}}$. We assume cluster-level variance is constant across all clusters.

## Assumption 4: Independence

We assume errors are uncorrelated with each other:

- $\color{#4daf4a}{e_{ij}}$ residuals are independent of other observations.
- $\color{#e41a1c}{u_j}$ deviations are independent across clusters<br>(e.g., School A's intercept doesn't inform School B's).
- $\color{#4daf4a}{e_{ij}}$ and $\color{#e41a1c}{u_j}$ are independent of each other.

## Assumption 4: The Exception

Are we allowed to have *any* correlations?

$$
\color{#e41a1c}{\Sigma} = \begin{bmatrix}\color{#e41a1c}{\tau_{00}} & \\ \color{#e41a1c}{\tau_{10}} & \color{#e41a1c}{\tau_{11}}\end{bmatrix}
$$

Yes! We explicitly model the correlation between a cluster's intercept and its slope via $\color{#e41a1c}{\tau_{10}}$. This "within-cluster" dependency is expected and accounted for by the model.


# Part 2:<br> Workflow \& Data

## Prepping the Data {.fs80}

```{r}
#| message: false
#| replace_path: ["../../data/", ""]

library(tidyverse)
library(glmmTMB)
library(easystats)

dat <- read_csv("../../data/diag_sim.csv") |> mutate(school = factor(school))
glimpse(dat)
```

## A Systematic Workflow

1. **Collinearity**: Check before you even look at residuals.
2. **Predictive Check**: Does the model "look" like the data?
3. **Residual Analysis**: L1 and L2 normality/variance.
4. **Sensitivity**: Does removing outliers change the story?

# Part 3:<br> Collinearity

## Redundant Predictors

Predictors that are too highly correlated with each other inflate Standard Errors, making "significant" effects disappear.

```{r}
# A model with redundant predictors
fit_coll <- glmmTMB(score_ideal ~ homework + study_time + (1 | school), data = dat)

# easystats check
check_collinearity(fit_coll)
```

## How to address collinearity?

- **Drop one**: If VIF is high, choose the variable that is more theoretically central.
- **Composite**: Average the variables into a single index.
- **MSEM**: Combine them into a latent variable (factor).

```{r}
# Creating a composite 'effort' score
dat <- dat |> mutate(effort = (homework + study_time)/2)

fit_fixed <- glmmTMB(score_ideal ~ effort + (1 | school), data = dat)
```

# Part 4:<br> Predictions

## Predictive Checks

- Before zooming in on residuals, we need to check the macroscopic fit of our model. 

- The `check_predictions()` function simulates new datasets based on your fitted model's parameters and compares their overall shape to your actual data. 

- **The Goal**: If your model accurately reflects the true data-generating process, the simulated predictions (the blue lines) should overlap your raw data (the green line).

## Plotting Predictions

```{r}
check_predictions(fit_fixed)
```

## Diagnosing Predictive Misfit {.fs80}

What if the model predictions do not align with the raw data? This indicates fundamental **misspecification**. No amount of residual tweaking will fix a model that cannot replicate the basic shape of your outcome.

- **Bimodal Data (Two humps):** The model is only predicting one peak, but the data has two. You are likely missing a critical categorical predictor or grouping variable that separates the observations.
- **Skewed Data (Long tails):** The model predicts a symmetrical curve, but the data has a long tail. You likely need to change the distribution family (e.g., moving to a Gamma or Poisson distribution) or apply a transformation.
- **Shifted Peaks:** The model's center is entirely offset from the data's center. You might be missing important non-linear effects, such as polynomial terms or interactions.

# Part 5: Residuals

## 1. Normality of Residuals

```{r}
#| eval: false

fit_skew <- glmmTMB(score_skew ~ homework + (1 + homework | school), data = dat)
check_normality(fit_skew, effects = "fixed") |> plot()
```

```{r}
#| echo: false
#| fig-width: 12
#| fig-height: 7

fit_skew <- glmmTMB(score_skew ~ homework + (1 + homework | school), data = dat)
check_normality(fit_skew, effects = "fixed") |> plot() + theme_gray(base_size = 24)
```

## Interpreting Residual Normality

- This plot checks the Level 1 errors ($\color{#4daf4a}{e_{ij}}$).
- The points should fall tightly along the green reference line.
- If points drift heavily at the tails, your residuals are skewed or heavy-tailed. Severe deviations here can bias your standard errors and $p$-values.

## 2. Normality of Random Effects

```{r}
#| eval: false

check_normality(fit_skew, effects = "random") |> plot()
```

```{r}
#| echo: false
#| fig-width: 12
#| fig-height: 7

p <- check_normality(fit_skew, effects = "random") |> plot()
p[[1]] + theme_gray(base_size = 20)
```

## Interpreting RE Normality

- This plot checks if the cluster-level deviations ($\color{#e41a1c}{u_j}$) follow a normal distribution.
- Each dot represents a specific cluster (e.g., a school).
- Extreme deviations here mean certain clusters behave completely differently than the rest of the population, which might require investigation.

## 3. Homogeneity of Variance

```{r}
#| eval: false

fit_het <- glmmTMB(score_het ~ homework + (1 + homework | school), data = dat)
check_heteroskedasticity(fit_het) |> plot()
```

```{r}
#| echo: false
#| fig-width: 12
#| fig-height: 5

fit_het <- glmmTMB(score_het ~ homework + (1 + homework | school), data = dat)
check_heteroscedasticity(fit_het) |> plot() + theme_gray(base_size = 20)
```

## Interpreting Homogeneity

- The green reference line should be relatively flat and horizontal.
- If the line bends sharply or the points fan out (like a megaphone), the variance is not constant ($\color{#4daf4a}{\sigma}$ is changing).
- In this plot, we can clearly see the variance expanding as the fitted values increase.

## Fix: Modeling Variance Directly

- If variance is not constant, we can actually model *why* it changes. The `glmmTMB` package allows us to predict the variance ($\sigma$) just like we predict the mean using the `dispformula` argument.

```{r}
#| eval: false
#| echo: true
# Modeling variance as a function of homework
fit_disp <- glmmTMB(
  score_het ~ homework + (1 | school),
  dispformula = ~ homework, 
  data = dat
)
```

*(Note: This creates a Location--Scale Mixed Model, where we estimate slopes for both the average and spread of the scores).*

# Part 6: Outliers

## Outliers in MLMs {.fs80}

- In standard regression, an outlier just pulls the line
- In MLMs, outliers distort our system of variances

- **L1 Outlier ($\color{#4daf4a}{e_{ij}}$)**: An observation that is extreme within its cluster.
  - *Impact*: Inflates the residual variance ($\color{#4daf4a}{\sigma}$), making all standard errors larger and reducing your power to detect true effects.
- **L2 Outlier ($\color{#e41a1c}{u_j}$)**: A cluster that is extreme among clusters.
  - *Impact*: Distorts the main fixed effects (slopes) and inflates cluster variance ($\color{#e41a1c}{\tau}$), making groups look more different than they actually are.

## Why DHARMa for Outliers? {.fs80}

- **The Problem with Traditional Outlier Metrics**: Tools like Cook's Distance struggle in MLMs because the model intentionally "shrinks" extreme cluster estimates toward the mean.
- Thus, `check_outliers()` does not yet support `glmmTMB` models.
- **The Simulation Solution**: `DHARMa` bypasses this by simulating thousands of new datasets directly from your fitted model parameters.
- **Flagging Outliers**: If a real data point falls *completely* outside the expected boundaries of those simulations, it is flagged as a structural outlier.

## Identifying Outliers

```{r}
#| echo: true
#| fig-show: hide
#| message: false

library(DHARMa)
fit_out <- glmmTMB(score_outlier ~ homework + (1 | school), data = dat)
res_out <- simulateResiduals(fit_out)
testOutliers(res_out)
```

## Plotting Outliers

```{r}
#| echo: false
#| results: hide

testOutliers(res_out)
```

## Interpreting DHARMa Outliers {.fs80}

- **The Histogram**: This plots your residuals as percentiles, not raw mathematical errors. Because a perfect model means a data point is equally likely to fall into any percentile, we want to see a flat, uniform block from 0.0 to 1.0, *not* a traditional bell curve.
- **The Red Bars**: DHARMa assigns a value of exactly 0 or 1 to any data point that falls completely outside the simulated boundaries. These extreme outliers are grouped and highlighted in red at the very edges.
- **The Title**: The title "Outlier test n.s." means the test is "not significant." Even though there is a red bar, the *number* of outliers is completely normal and expected for a dataset of this size.

## Fix: Sensitivity Analysis

There is no magical R function to automatically "fix" outliers. You must use logic.

1.  Identify the outliers using DHARMa.
2.  Run the model **with** them.
3.  Run the model **without** them.
4.  **Compare**: If the main slope changes drastically, the outlier is "influential" and the results from both models should be reported transparently.

## Performing the Sensitivity Analysis

To run the model without outliers, we first need to extract them. DHARMa flags outliers by assigning them a simulated residual of exactly `0` or `1`.

```{r}
#| echo: true
# 1. Get the row numbers of the outliers
outlier_idx <- which(res_out$scaledResiduals %in% c(0, 1))

# 2. Create a clean dataset by dropping those rows
dat_clean <- dat |> slice(-outlier_idx)

# 3. Re-run the model on the clean data
fit_clean <- glmmTMB(score_outlier ~ homework + (1 | school), data = dat_clean)
```

## Comparing the Results

Now we use `compare_parameters()` from the `easystats` package to look at both models side-by-side. 

```{r}
compare_parameters(fit_out, fit_clean, select = "{estimate} ({p})")
```

## Interpreting the Comparison

*Notice what happened to our slope for `homework`.*

1.  **With Outliers**: The effect of homework was severely underestimated (e.g., heavily pulled by the extreme points).
2.  **Without Outliers**: The effect of homework recovered to its true value (closer to our simulated 1.2).
3.  **Conclusion**: These outliers are highly influential. You should report the results of the clean model but be completely transparent in your manuscript about how many outliers were removed and why.
