---
lecture: "14a"
metadata-files:
  - ../../_slide-settings.yml
format:
  revealjs:
    theme: default
---

::: {.my-title}
# [Multilevel Modeling]{.blue}
Effect Sizes & Reporting

::: {.my-grey}
[{{< meta semester >}} | CLAS | {{< meta course >}}]{}<br />
[Jeffrey M. Girard | Lecture {{< meta lecture >}}]{}
:::

![](../../img/city-girl.svg){.absolute bottom=30 right=0 width="400"}
:::

```{r setup}
#| echo: false
#| message: false

library(tidyverse)
library(easystats)
library(patchwork)
library(here)
library(glmmTMB)

source(here("_common.R"))

set_theme(theme_classic(base_size = 20))

options(width = 75)
```

## Roadmap

::: {.columns .pv4}

::: {.column width="60%"}
1. Effect Sizes
    + *Marginal vs. Conditional R²*
    + *StdX and StdYX Effects*
    
2. The Reporting Checklist
    + *Intro and Methods*
    + *Results and Discussion*
    
:::

::: {.column .tc .pv4 width="40%"}
{{< lif "../../icons/map.json" trigger=hover colors=secondary:#2a76dd class=rc >}}
:::

:::

# Effect Sizes in MLMs

## Beyond *p*-Values {.fsmaller}

:::{.columns}
:::{.column width="60%"}
- Statistical significance tells us if an effect exists. It does not tell us if the effect is meaningful.
- **Effect sizes** contextualize the magnitude of results from your model.
- They answer the question: "What is the practical significance of this result?"
- Reviewers and journal editors increasingly demand effect sizes for all key findings.
:::
:::{.column .tc .pv4 width="40%"}
{{< lif "../../icons/rulers.json" trigger=hover colors=secondary:#2a76dd class=rc >}}
:::
:::

## Two Types of Effect Sizes {.fsmaller}

**Unstandardized Effect Sizes**

- Expressed in the raw units of the dependent variable
- *e.g., For every extra year of service, salary increases by $1000*
- Useful when the outcome variable has an intuitive metric

**Standardized Effect Sizes**

- Measure magnitude without units
- *e.g., The model explains 24 percent of the variance in patient satisfaction*
- Useful when the original metrics are abstract

## The R-squared Metric {.fsmaller}

- The most common standardized effect size is R²
- It represents the proportion of variance in the outcome variable explained by the model and all its included effects
- In single-level regression, the math is straightforward:
$$R^2 = \frac{\text{Explained Variance}}{\text{Total Variance}}$$
- It yields an intuitive measure ranging from:<br>
0 (nothing explained) to 1 (everything explained)
- However, applying this to Multilevel Models is notoriously difficult

## The Multilevel Variance Problem {.fsmaller}

Why can't we just use a single R² for an MLM?

- **Multiple Denominators:**<br>
Total variance is partitioned. Do we want to explain total outcome variance, within-cluster variance, or between-cluster variance?
- **Multiple Numerators:**<br>
We have multiple sources of explained variance. We have fixed effects at Level-1, fixed effects at Level-2, random intercepts, and random slopes.
- Summing these up into a single neat number obscures the complexity of the data structure and can be misleading.

## The Rights & Sterba Framework {.fsmaller}

- Rights \& Sterba (2019) developed a complex framework for R² in MLMs.
- Because there are so many combinations of numerators and denominators, their framework yields **12 different R-squared measures** for a single model.
- For example:
  - Proportion of total variance explained by Level-1 fixed effects.
  - Proportion of within-cluster variance explained by random slopes.
  - Proportion of between-cluster variance explained by random intercepts.

## Practical Constraints {.fsmaller}

- This framework is academically rigorous and statistically exact.
- If your research question relies on isolating exactly how much Level-2 variance is explained by a Level-1 random slope, it is the tool to use.
- However, for most applied researchers, reporting 12 effect sizes per model is overwhelming for both the author and the reader.
- Most journals expect a simpler, two-part summary of model performance.

## The Nakagawa Framework {.fsmaller}

- The widely accepted standard for applied MLMs is the framework developed by Nakagawa \& Schielzeth (2013).
- It simplifies the problem by providing two metrics:
  - **Marginal R²:** Variance explained by only the fixed effects.
  - **Conditional R²:** Variance explained by the entire model<br>(fixed effects plus random effects).
- This is what we will usually report in practice.

# Calculating<br> R-squared

## Setup and Data

```{r}
#| message: false

library(tidyverse)
library(glmmTMB)
library(easystats)
```

```{r}
#| message: false
#| replace_path: ["../../data/", ""]

hospital <- read_csv("../../data/hospital_sim.csv")
glimpse(hospital)
```

## Preparing the Data

:::{.fsmaller}
To calculate accurate R-squared values, we do not need to standardize or center our Level-2 predictors. However, we **must** decompose our Level-1 predictors to properly isolate the within-cluster and between-cluster variance.
:::

```{r}
l2_means <- hospital |> 
  summarize(.by = hospital, wait_b = mean(wait_time))

dat_unstd <- hospital |> 
  left_join(l2_means, by = "hospital") |> 
  mutate(wait_w = wait_time - wait_b)
```

## Fitting the Baseline Model

:::{.fsmaller}
Now we fit our model using our unstandardized, centered predictors.
:::

```{r}
fit_unstd <- glmmTMB(
  formula = satisfaction ~ 1 + wait_b + wait_w + nurse_ratio + 
    (1 + wait_w | hospital), 
  data = dat_unstd,
  REML = TRUE
)
```

## Extracting R-squared

:::{.fsmaller}
We can extract our Nakagawa R² values cleanly using the \{performance\} package from the \{easystats\} ecosystem.
:::

```{r}
r2(fit_unstd)
```

## Interpreting R-squared {.fsmaller}

- **Marginal R² (0.176):** Fixed effects explain ~17.6% of the variance.
  - *Prediction Framing:* How well we predict a new observation from a **new, unseen cluster**. We must rely solely on the population averages.
- **Conditional R² (0.642):** The entire model (fixed + random) explains ~64.2% of the variance.
  - *Prediction Framing:* How well we predict a new observation from an **existing, known cluster**. We can fine-tune the prediction using that cluster's random effects.
- **The Difference:** The random effects alone account for roughly 46.6% of the variance (Conditional minus Marginal).

# Standardized Effects

## To Standardize Y or Not? 

:::{.fsmaller}
Before we standardize our slopes, we must ask a theoretical question about our outcome variable (Y).

- **Meaningful Outcomes:** If Y has a well understood and intuitive metric (e.g., dollars, days, blood pressure), we leave it in its raw units. We only standardize the predictors, yielding [StdX]{.b .blue} slopes.
- **Arbitrary Outcomes:** If Y is an abstract scale (e.g., a satisfaction survey, an anxiety inventory), raw points are less intuitive. We standardize the predictors AND the outcome, yielding [StdYX]{.b .blue} slopes.
:::

## The Predictor Trap 

:::{.fsmaller}
Regardless of using StdX or StdYX, we must be incredibly careful about how we standardize our multilevel predictors (X).

- **The "Global" Trap:** Standard software defaults to standardizing based on the total sample's mean and SD. Here is why that fails:
    + **Mixing Levels:** A Level-1 predictor can only explain within-cluster variance. Dividing its slope by the total SD (which includes between-cluster differences) unfairly penalizes it, making strong local effects look artificially weak.
    + **Cluster Weighting:** Level-2 predictors are attributes of the cluster. If you standardize across all data rows, a cluster with 500 observations influences the math 10x more than a cluster with 50 observations, distorting the meaning of a "standard unit" for that variable.
    
:::

## StdX: Data Preparation (L2)

:::{.fsmaller}
To avoid mixing levels, we create a distinct Level-2 dataset using summarize. Standardizing within this dataset uses the correct means and SDs.
:::

```{r}
# 1. Create a distinct Level-2 dataset
l2_std <- hospital |> 
  summarize(
    wait_b = mean(wait_time),
    nurse_ratio = first(nurse_ratio),
    .by = hospital
  ) |> 
  mutate(
    # 2. Standardize across the Level-2 clusters
    wait_b_z = standardize(wait_b),
    nurse_z = standardize(nurse_ratio)
  )
```

## StdX: Data Preparation (L1)

:::{.fsmaller}
Next, we join those standardized L2 variables back to our L1 data and strictly standardize the within-cluster variance.
:::

```{r}
dat_stdx <- hospital |> 
  # 3. Join the Level-2 data back to the main dataset
  left_join(l2_std, by = "hospital") |> 
  mutate(
    # 4. Isolate and standardize the within-cluster variance
    wait_w = wait_time - wait_b,
    wait_w_z = standardize(wait_w)
  )
```

## StdX: Model & Interpretation

```{r}
fit_stdx <- glmmTMB(
  formula = satisfaction ~ 1 + wait_b_z + wait_w_z + nurse_z + 
    (1 + wait_w_z | hospital), 
  data = dat_stdx,
  REML = TRUE
)
model_parameters(fit_stdx, effects = "fixed")
```

:::{.fsmaller}
- A one *SD* increase in a patient's wait time (relative to their hospital's average) significantly decreases their satisfaction by 0.37 *raw points*.
:::

## The Problem with Global StdYX {.fsmaller}

If Y is an arbitrary scale, you might be tempted to standardize Y globally before fitting your model. [Do not do this.]{.b .dark-red} 

- **The ICC Penalty:** If you standardize Y globally, the denominator includes both Level-1 and Level-2 variance. 
- **The Trap:** A Level-1 predictor can only explain Level-1 variance. By dividing its effect by the *total* variance pool, you mathematically penalize it. 
- **The Result:** If your model has high Level-2 variance (high ICC), your Level-1 StdYX effects will look artificially tiny.
- **The Solution:** We must use [Level-Specific Standardization]{.b .blue}.

## Level-Specific StdYX 

:::{.fsmaller}
There is no automated R package for this (yet). We manually take the coefficients from our **StdX model** (where X is standardized but Y is raw) and divide them by the level-specific standard deviations of Y.
:::

```{r}
# 1. Calculate the standard deviation of the hospital means (Level 2 SD)
sd_y_b <- hospital |> 
  summarize(sat_b = mean(satisfaction), .by = hospital) |> 
  pull(sat_b) |> sd() |> print()

# 2. Calculate the standard deviation of the within-cluster scores (Level 1 SD)
sd_y_w <- hospital |> 
  mutate(sat_b = mean(satisfaction), .by = hospital) |> 
  mutate(sat_w = satisfaction - sat_b) |> 
  pull(sat_w) |> sd() |> print()
```

## Calculating & Interpreting StdYX

:::{.fsmaller}
We divide the slopes from our `fit_stdx` model by the matching SD of Y.
:::

```{r}
# L2 StdYX Slope = (L2 StdX Slope) / (L2 SD of Y)
0.19 / 0.700 # nurse_z
```

:::{.fsmaller}
- A one *between-hospital* SD increase in nurse ratio was associated with a non-significant increase in satisfaction of 0.27 *between-hospital* SDs.
:::

```{r}
# L1 StdYX Slope = (L1 StdX Slope) / (L1 SD of Y)
-0.37 / 0.722 # wait_w_z
``` 

:::{.fsmaller}
- A one *within-hospital* SD increase in wait time was associated with a significant decrease in satisfaction of 0.51 *within-hospital* SDs.
:::

# The Reporting<br> Checklist

## The Goal of Reporting {.fsmaller}

- Multilevel modeling involves dozens of minor analytical decisions. 
- Centering, random effect structures, and estimation algorithms all drastically change your results.
- Your goal in a manuscript is not just to report the p-values, but to provide enough transparent detail that another researcher could perfectly replicate your model from your raw data.

## IMRD: Introduction {.fsmaller}

The justification for an MLM should begin before the Methods section.

**What to include:**

- Acknowledge the nested structure of your data in your hypotheses.
- Clearly state the levels of analysis (e.g., L1 is patients, L2 is hospitals).
- Distinguish between within-cluster and between-cluster hypotheses. Do not hypothesize a generic "effect of wait time" if you specifically mean "when a patient waits longer than their hospital's average."

## IMRD: Methods (Data) {.fsmaller}

The participants section must describe the sample sizes at all levels.

**What to include:**

- Total number of Level-1 observations (e.g., N = 2,000 patients).
- Total number of Level-2 clusters (e.g., J = 40 hospitals).
- The average cluster size, minimum cluster size, and maximum cluster size (e.g., an average of 50 patients per hospital, ranging from 12 to 89).

## IMRD: Methods (Variables) {.fsmaller}

You must explicitly state how every variable was handled.

**What to include:**

- Identify which variables were measured at Level-1 and Level-2.
- Explicitly state your centering decisions. Did you grand-mean center? Did you group-mean center (cluster-mean center)? 
- State why you chose that centering method. Did you want to isolate the purely within-cluster effect, or were you aiming for a blended effect?

## IMRD: Methods (Modeling) {.fsmaller}

Clearly identify the mathematical tools you used.

**What to include:**

- Software and packages with version numbers (e.g., R version 4.3.1, glmmTMB version 1.1.8).
- The estimation method used. Did you use Restricted Maximum Likelihood (REML) or standard Maximum Likelihood (ML)? 
- Mention the specific optimizer if you had to change it from the default to achieve convergence.

## IMRD: Methods (Structure) {.fsmaller}

Explain how you built your equations.

**What to include:**

- Describe the random effects structure. Did you include random intercepts? Random slopes?
- State whether you allowed the random intercepts and slopes to correlate.
- If you had to simplify your random effects structure due to singular fits or convergence issues, document that process here. Explain what was dropped and why.

## IMRD: Methods (Descriptives & Reliability) {.fsmaller}

You cannot simply present a correlation matrix for a multilevel dataset.

**What to include:**

- A table of means, standard deviations, and correlations for all Level-1 variables.
- A completely separate table of descriptives and correlations for all Level-2 variables, calculated at the cluster level.
- Evidence of reliability for your continuous predictors. Predictors measured with high error will bias your regression coefficients.

## IMRD: Methods (Missing Data) {.fsmaller}

Missing data is especially dangerous in MLMs. Deleting one Level-2 cluster due to a missing covariate takes all of its Level-1 observations down with it.

**What to include:**

- The exact percentage of missing data reported separately for Level-1 variables and Level-2 variables.
- The method used to handle missingness (e.g., Full Information Maximum Likelihood or Multiple Imputation).
- If using listwise deletion, explicitly state the final analytical sample size and justify why deletion was acceptable.

## IMRD: Results (The Null Model) {.fsmaller}

Always report the baseline before adding predictors.

**What to include:**

- Report the unconditional ICC from the null model.
- Provide a plain language interpretation: "The ICC was 0.22, indicating that 22 percent of the variance in patient satisfaction is attributable to differences between hospitals."
- This justifies the need for multilevel modeling in the first place.

## IMRD: Results (Fixed Effects) {.fsmaller}

Report the focal tests of your hypotheses.

**What to include:**

- The unstandardized coefficient estimate.
- The 95% Confidence Interval for the estimate.
- The standard error.
- The exact p-value.
- A plain language interpretation of the direction and magnitude of the effect.

## IMRD: Results (Random Effects) {.fsmaller}

Do not ignore the random portion of the model.

**What to include:**

- The variance (or standard deviation) of the random intercepts.
- The variance (or standard deviation) of any random slopes.
- The correlation between random intercepts and random slopes.
- The residual (Level-1) variance.

## IMRD: Results (Effect Sizes) {.fsmaller}

Provide the big picture summary of model performance.

**What to include:**

- Report both the Marginal R2 and Conditional R2.
- Cite Nakagawa \& Schielzeth (2013) to clarify your framework.
- If comparing multiple models (e.g., a baseline model vs a full model), report the change in R2 (Delta R2) or use AIC/BIC to discuss relative model fit.
- Clearly state whether you are reporting raw, StdX, or StdYX slopes.

## IMRD: Results (Model Building Sequence) {.fsmaller}

Reviewers want to see the sequence of models you estimated.

**What to include:**

- **Step 1:** The Null Model to establish baseline variance and the ICC.
- **Step 2:** The Random Coefficients Model containing only Level-1 predictors to see if those slopes vary significantly across clusters.
- **Step 3:** The Full Contextual Model containing Level-1 and Level-2 predictors, including any cross-level interactions.

## IMRD: Discussion (Language) {.fsmaller}

Multilevel modeling uses the word "effect" constantly (e.g., fixed effects, random effects). This can create a semantic trap.

**What to watch out for:**

- Do not let the statistical terminology bleed into causal claims in your discussion section.
- Unless you have a rigorously controlled experimental design, your "fixed effects" are merely associations or predictions.
- Carefully edit your conclusions to ensure you are not implying a causal mechanism that your design cannot support.
