---
format: 
  revealjs:
    css: ../../styles.css
    slide-number: true
    show-slide-number: all
    preview-links: false
    progress: true
    history: true
    hash-type: number
    theme: default
    code-block-background: true
    highlight-style: zenburn
    code-link: false
    code-copy: true
    code-line-numbers: false
    controls: true
    pagetitle: "Multilevel Modeling"
    author-meta: "Jeffrey Girard"
    semester: "Spring 2026"
    course: "PSYC 894"
    lecture: "03a"
execute:
  echo: true
  eval: true
  collapse: false
  cache: false
---

::: {.my-title}
# [Multilevel Modeling]{.blue}
Issues with Clustered Data

::: {.my-grey}
[{{< meta semester >}} | CLAS | {{< meta course >}}]{}<br />
[Jeffrey M. Girard | Lecture {{< meta lecture >}}]{}
:::

![](../../img/city-girl.svg){.absolute bottom=30 right=0 width="400"}
:::

## Roadmap

::: {.columns .pv4}

::: {.column width="60%"}
1. Multilevel Data and Questions
  
2. Conceptual Issues with Clustering

3. Statistical Issues with Clustering
:::

::: {.column .tc .pv4 width="40%"}
{{< lif "../../icons/map.json" trigger=hover colors=secondary:#2a76dd class=rc >}}
:::

:::

##  R Setup

```{r}
#| message: false

library(easystats)
library(ggplot2)
library(marginaleffects)

data("penguins")
```

:::{.fsmaller}
- [easystats]{.b .blue} will provide most of our statistical functions
- [ggplot2]{.b .blue} will provide the ability to customize our plots
- [marginaleffects]{.b .green} is called by easystats for estimation
:::

## Example Dataset

```{r}
# This penguins dataset is built into R (don't load palmerpenguins)
data_codebook(penguins, 
  select = c("species", "bill_dep", "flipper_len", "body_mass"))
```

